Delayed effects of [responses to?] climate lead to demographic divergence in a fragmented Amazonian landscape

### 2021-04-26 doc/paper.Rmd
more discussion edits - looking good

Populations in fragmented tropical landscapes are also at risk from climate change [@brodieClimateChangeTropical2012; @scheffersExtremeThermalHeterogeneity2017], which can have complex or even contrasting effects across the same tropical biome.
For instance, <!--#parts of the Amazon basin have experienced a drying trend since the mid-1970's, and --> climate models for the next 50-100 years predict that some parts of Amazonia will experience decreases in dry season precipitation [@malhiClimateChangeDeforestation2008] and increases in the frequency and severity of droughts [@caiIncreasingFrequencyExtreme2014; @zengCausesImpacts20052008; @moraProjectedTimingClimate2013].
These changes could exacerbate the altered environmental conditions in fragments thought to make populations demographically vulnerable - particularly for plants, which are unable to move in response to changes in local conditions.
In contrast, other parts of the basin are predicted to see an increase in the frequency and severity of extreme precipitation events, which can also have detrimental consequences for plant survival and reproduction [@estebanOtherSideDroughts2021].

@@

Plants in continuous forest had an average of 2.9 shoots (± 1.8 SD) and were on average 40.6 cm tall (± 26.5 SD).
Plants in 1-ha fragments had on average **\~XX% fewer** shoots (2.5 ± 1.5 SD) and **were XX% shorter** (36.2 cm ± 24.1 SD).
Because the proxy for plant size used in our models was the product of these two metrics, plants in fragments were consequently significantly larger than those in forest fragments (150 ± 175 SD vs. 112 ± 141 SD, respectively).
This disparity in plant size was most notable in the initial years of our surveys and then diminished over time (Figure \@ref(fig:eda)a), likely due to a combination of greater seedling recruitment in CF and plant mortality or shrinkage As with survival, size in year t was a significant predictor of size in year t+1 (p \< 0.001).
While the effect was generally similar across size classes and habitat types, the impact of plant size on growth was greatest for mid-sized plants in continuous forest (Figure \@ref(fig:size-plot)b).

@@

Understanding how landscape structure and environmental conditions act to influence population dynamics is fundamental to many of the conceptual frameworks for studying and conserving fragmented landscapes [@didhamRethinkingConceptualFoundations2012; @driscollConceptualDomainMatrix2013].
<!--# more here -->.
Despite this, surprisingly few studies have compared the structure and demography of populations in habitat fragments and continuous habitat; those that have are typically short term (i.e., \<4 years).
Using a decadal dataset of plant demography, we show that some - but not all - results from short-term studies can be extrapolated to longer times scales.
Our results also add to the emerging body of work showing delayed effects of climate drivers on demographic vital rates [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
Most Finally, we show for the first time that the magnitude, direction, and temporal lag in demographic responses to drought differs between forest fragments and continuous forest.

@@

Decadal comparisons of *Heliconia acuminata* growth and survival largely mirrored those of prior experiments and surveys [@brunaHabitatFragmentationDemographic2002].
In contrast, conclusions regarding the effect of fragmentation on flowering differed substantially from those based on short-term surveys [@brunaHabitatFragmentationDemographic2002].
While in some years the probability a plant flowered was far lower in fragments than in continuous forest, in others it was comparable or even higher in fragments.
The probability that individual plants flowered continued to be size-dependent, with a similar size threshold below which plants were unlikely to produce inflorescences.
However, while in our initial three-year survey the relationship between plant size and probability of flowering was similar in both habitat types [@brunaHabitatFragmentationDemographic2002], when using the decadal dataset were found that flowering probability for the largest plants is actually higher in forest fragments (i.e., average fitted probability of flowering for plants in upper size quartile = 0.125 in CF vs. 0.094 in FF).
These size-dependent differences in flowering probability had only a modest effect on the previously observed disparities in flowering plant abundance - populations in fragments are smaller, and they have proportionately fewer of these large plants [@brunaHabitatFragmentationDemographic2002].
However, these differences in flowering probability could potentially lead to major habitat-specific differences in lifetime reproductive effort for these long-lived plants, as they translate into flowering once every 7.97 years in fragments vs. every 10.63 years in continuous forest.
Our results underscore why caution is needed when using short-term studies to draw conclusions regarding the effects of fragmentation on long-lived organisms: seemingly small (i.e., non-significant) differences in vital rates, when compounded over multiple years, can lead to demographic divergence among populations [@gagnonGrowthUnderstoryHerb2011].

@@

This study adds to the growing body of work that finds delayed effects of climate drivers on vital rates without *a priori* assumptions about the influence of particular climate windows [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
We found evidence for delayed effects of drought which are sometimes stronger predictors of vital rates than more recent conditions.
The approach we took offers an unbiased way of identifying these delayed effects without overfitting (but see @pierreCriticalWindowsMethod2020 and @ogleQuantifyingEcologicalMemory2015 for alternative methods), yet few demographic studies have the data required for modeling lagged effects [@tellerLinkingDemographyDrivers2016; @croneHowPlantEcologists2011].
This demonstrates the importance of long term demographic experiments in determining the effects of changing habitats on populations.
<!--# Or something like "some of these long-term studies exist and are prime for doing these kinds of analyses" -->

@@

We found that the effects of variation in SPEI differed between forest fragments and continuous forest habitat.
At mean SPEI, overall survival of plants was similar in fragments and continuous forest, but extremes in SPEI, both positive (wet) and negative (dry) had much greater detrimental effects on survival in habitat fragments.
Similarly variation in SPEI had a greater impact on plant size in fragments compared to continuous forest with a general trend of unusually wet wet seasons and unusually dry dry seasons having a greater negative impact in fragments.
Flowering probability showed a different pattern with a greater effect of variation in SPEI in continuous forests.
However, with few plants large enough to flower, it is likely we simply did not detect the full effects of variation in SPEI on flowering probability in forest fragments with our models.
Both variation in precipitation and extreme conditions are predicted to increase as climate change progresses ( <!--# cite IPCC? -->) and our results demonstrate that this may have a greater negative impact on understory plants in forest fragments compared to continuous forests.

@@

-   Very few demographic studies have the data required to fit models that include delayed effects.

>>>>

Populations in fragmented tropical landscapes are also at risk from climate change [@brodieClimateChangeTropical2012; @scheffersExtremeThermalHeterogeneity2017], which can have complex or even contrasting effects across the same tropical biome. For instance, <!--#parts of th      e Amazon basin have experienced a drying trend since the mid-1970's, and --> climate models for the next 50-100 years predict that some parts of Amazonia will experience decreases in dry season precipitation [@malhiClimateChangeDeforestation2008] and increases in the frequency and severity of droughts [@caiIncreasingFrequencyExtreme2014; @zengCausesImpacts20052008; @moraProjectedTimingClimate2013]. These changes could exacerbate the altered environmental conditions in fragments thought to make populations demographically vulnerable - particularly for plants, which are unable to move in response to changes in local conditions. In contrast, other parts of the basin are predicted to see an increase in the frequency and severity of extreme precipitation events, which can also have detrimental consequences for plant survival and reproduction [@estebanOtherSideDroughts2021].

@@

Plants in continuous forest had an average of 2.9 shoots (± 1.8 SD) and were on average 40.6 cm tall (± 26.5 SD).
Plants in 1-ha fragments had on average **\~XX% fewer** shoots (2.5 ± 1.5 SD) and **were XX% shorter** (36.2 cm ± 24.1 SD).
Because the proxy for plant size used in our models was the product of these two metrics, plants in fragments were consequently significantly larger than those in forest fragments (150 ± 175 SD vs. 112 ± 141 SD, respectively).
This disparity in plant size was most pronounced in the initial years of our surveys and then diminished over time (Figure \@ref(fig:eda)a), likely due to a combination of greater seedling recruitment in CF and plant mortality or shrinkage As with survival, size in year t was a significant predictor of size in year t+1 (p \< 0.001).
While the effect was generally similar across size classes and habitat types, the impact of plant size on growth was greatest for mid-sized plants in continuous forest (Figure \@ref(fig:size-plot)b).

@@

Understanding how landscape structure and environmental conditions act to influence population dynamics is fundamental to many of the conceptual frameworks for studying and conserving fragmented landscapes [@didhamRethinkingConceptualFoundations2012; @driscollConceptualDomainMatrix2013]. <!--# more here -->. Despite this, surprisingly few studies have compared the structure and demography of populations in habitat fragments and continuous habitat; those that have are typically short term (i.e., \<4 years). Using a decadal dataset of plant demography, we show that some - but not all - results from short-term studies can be extrapolated to longer times scales. Our results also add to the emerging body of work showing delayed effects of climate drivers on demographic vital rates [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
Finally, we show for the first time that the magnitude, direction, and temporal lag in demographic responses to drought differs between forest fragments and continuous forest.

Understanding how landscape structure and environmental conditions act to influence population dynamics is fundamental to many of the conceptual frameworks for studying and conserving fragmented landscapes [@didhamRethinkingConceptualFoundations2012; @driscollConceptualDomainMatrix2013].
Using a decadal dataset of plant demography, we show for the first time that the magnitude, direction, and temporal lag in demographic responses to drought differs between forest fragments and continuous forest.

@@

Decadal comparisons of *Heliconia acuminata* growth and survival largely mirrored those of prior experiments and surveys [@brunaHabitatFragmentationDemographic2002]. In contrast, conclusions regarding the effect of fragmentation on flowering differed substantially from those based on short-term surveys [@brunaHabitatFragmentationDemographic2002]. While in some years the probability a plant flowered was far lower in fragments than in continuous forest, in others it was comparable or even higher. The probability that individual plants flowered continued to be size-dependent, with a similar size threshold below which plants were unlikely to produce inflorescences. However, while in our initial three-year survey the relationship between plant size and probability of flowering was similar in both habitat types [@brunaHabitatFragmentationDemographic2002], when using the decadal dataset were found that flowering probability for the largest plants is actually higher in forest fragments (i.e., average fitted probability of flowering for plants in upper size quartile = 0.125 in CF vs. 0.094 in FF). These size-dependent differences in flowering probability had only a modest effect on the previously observed disparities in flowering plant abundance - populations in fragments are smaller, and they have proportionately fewer of these large plants [@brunaHabitatFragmentationDemographic2002]. However, these differences in flowering probability could potentially lead to major habitat-specific differences in lifetime reproductive effort for these long-lived plants, as they translate into flowering once every 7.97 years in fragments vs. every 10.63 years in continuous forest. Our results underscore why caution is needed when using short-term studies to draw conclusions regarding the effects of fragmentation on long-lived organisms: seemingly small (i.e., non-significant) differences in vital rates, when compounded over multiple years, can lead to demographic divergence among populations [@gagnonGrowthUnderstoryHerb2011].

@@

Precipitation and other climatic factors are known to have large effects on plant growth, survival, or reproduction.
Despite this, models of population dynamics have rarely included the effects of environmental drivers.
This has largely due to the challenge - both ecologically and statistically - of detecting any demographic responses to climatic extremes that are delayed for multiple growing seasons.
To address this, researchers have begun to use a number of statistical methods that test for time lags in demographic responses without *a priori* assumptions about the influence of any particular climate window [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021]. Our expansion of this approach, which offers an unbiased way of identifying these delayed effects without overfitting (but see @pierreCriticalWindowsMethod2020 and @ogleQuantifyingEcologicalMemory2015 for alternative methods) yielded results consistent with this emerging literature -- that the effects of precipitation extremes on the demography of *Heliconia acuminata* could be delayed for up to **XXX growing seasons**.

However, while it appeared that delayed effects of climate on demographic vital rates were ubiquitous, the extent to which they vary spatially or with habitat remained an open question [@eversLaggedDormantSeason2021]. Our results clearly indicate that they can, with habitat-specific differences in how environmental conditions influenced future vital rates. For example, [delete this phrase? the overall survival of plants was similar in continuous forest and forest fragments, and] although extreme values of SPEI - both positive (unusually high precipitation) and negative (drought conditions) - led to declines in the probability of individual survival both habitat types.
However, the magnitude of these declines was far greater in forest fragments.
Similarly, the detrimental effects on growth rates of variation in SPEI, in this case unusually rainy wet seasons and dry seasons with unusually low precipitation, were also more severe in fragments.
In contrast, variation in SPEI had a far stronger effect on populations in continuous forests, though it is is important to recognize that because there are fewer plants in fragments above the threshold-size for flowering it is possible we lacked the statistical power to detect the full effects of SPEI in fragments.

Both variation in precipitation and extreme conditions are predicted to increase as climate change progresses ( <!--# cite IPCC? -->) and our results demonstrate that this may have a greater negative impact on understory plants in forest fragments compared to continuous forests.

@@

@@

-    few demographic studies have the data required for modeling lagged effects [@tellerLinkingDemographyDrivers2016; @croneHowPlantEcologists2011].
    This demonstrates the importance of long term demographic experiments in determining the effects of changing habitats on populations.
    <!--# Or something like "some of these long-term studies exist and are prime for doing these kinds of analyses" -->

-
